---
title: "Results [needs to be fine-tuned]"
format: 
    html:
       theme: journal
       self-contained: true
editor: visual
bibliography: references.bib
---

In this section, I will summarize study population characteristics and findings from analysis.

## Study I

### Descriptive

In The Maastricht Study, \[10,000 participated by Date\], of those 1316 reported prior CVD. Participants who had valid 24-hour HRV measured was 4379 and of those 3673 had a valid measurement of either CD or PWV. Study population included 3673 participants. Further characteristic are described in the study I manuscript \[Table 1\] \[refeernce to study I\].

```{r}
#| echo: false
#| warning: false
#| message: false
library(vroom)
library(here)
library(dplyr)

table_mas <- vroom(here("tables/table1_diabetes_status.csv"))

table_mas %>%
  dplyr::filter(!`**Characteristic**` %in% c("Ethnicity",
                                             "White",
                                             "Non-white",
                                             "Education",
                                             "Low (No education, (un)completed primary education, or lower vocational education)",
                                             "Middle (intermediate vocational education or higher secondary education)",
                                             "High (Higher vocational education or university education)",
                                             "None", 
                                             "Alcohol consumption",
                                             "Low (Women: ≤ 7, Men: ≤ 14)", 
                                             "High (Women: > 7, Men: > 14)",
                                             "Low (Women: â‰¤ 7, Men: â‰¤ 14)",
                                             "Smoking status", "Never", 
                                             "Former (quit > 6 months ago)", 
                                             "Former (quit < 6 months ago)", 
                                             "Current",  
                                             "ECGvar_LFHF", "ECGvar_NN50", 
                                             "ECGvar_SDSD", "med_DM")) %>%
      mutate_all(~ ifelse(is.na(.), "", .)) %>% 
  knitr::kable() %>%
  kableExtra::kable_classic() %>%
  kableExtra::kable_styling(latex_options = "scale_down")
```

### 24-hour HRV and arterial stiffness

**Time-domain HRV**

In the fully adjusted model 2, PWV was 2.8% (CI: 2.1; 3.4) lower, while CD was 3.3% (CI: 1.5; 5.1) higher per SD increase in HRV time-domain Z-score. Among the time-domain indices, SDNN, SDNNi, and SDANN showed the strongest associations, with cf-PWV being lower by 2.5% (CI: 2.0; 3.1), 2.5% (CI: 1.9; 3.4), and 2.2% (CI: 1.7; 2.7), respectively. Conversely, CD was higher by 3.2% (CI: 1.7; 4.7), 3.0 % (CI: 1.4; 4.6), and 2.8% (CI: 1.3; 4.3), respectively. RMSSD and pNN50 showed a weaker association with cf-PWV (-1.1% \[CI: -1.4; -0.4\], and -1.1 \[-1.7; -0.6\]), while no evidence for an association was found with CD.

**Frequency-domain HRV**

In the fully adjusted model 2, PWV was 2.8% (CI: 2.1; 3.5) lower, while CD was 3.2% (CI: 1.3; 5.1) higher per SD increase in HRV frequency-domain Z-score. Among the frequency-domain indices, total power, VLF, and ULF showed the strongest associations, with cf-PWV being lower by 2.2% (CI: 1.7; 2.8 ), 2.4% (CI: 1.9; 4.0), and 2.1% (CI: 1.5; 2.6), respectively. Conversely, CD was higher by 2.7% (CI: 1.2; 4.2), 2.4% (CI: 0.9; 4.1), and 2.6% (CI: 1.1; 4.1), respectively. HF showed a weaker association with cf-PWV (-0.9% \[CI: -1.4; -0.4\]), while no evidence for an association was found with CD. Mean interbeat interval was associated with 2.4 % (CI: 1.8; 2.9) lower cf-PWV and 4.5% (3.1; 6.1) higher CD.

### Effect modification of diabetes status

The study population represented diabetes risk of normal glucose metabolism (65%), prediabetes (15%), and type 2 diabetes (20%). The median (IQR) cf-PWV (aortic stiffness) increased with diabetes status: NGM: 8.08 m/s (7.28, 9.16), prediabetes: 8.96 m/s (7.84, 10.32), and type 2 diabetes: 9.36 m/s (8.16, 10.80). CD (carotid stiffness) decreased: NGM: 15.0 (11.8, 18.8), prediabetes: 13.5 (10.4, 16.9), and type 2 diabetes: 12.5 (9.9, 16.0) × 10⁻³/kPa. SDNN (ms) was highest in NGM and decreased with worsening glucose metabolism: NGM: 138ms (117, 164), prediabetes: 127ms (106, 152), and type 2 diabetes: 116ms (96, 139).

The association between HRV time-domain Z-scores and cf-PWV and CD was significantly modified by prediabetes (PWV: -4.9 \[CI: -6.523; -3.243\] $^{interaction(*) ^{p-value< 0.01}}$ CD: 8.0 \[CI:3.8; 12.5\]$^{*^{p-value< 0.01}}$) but not by type 2 diabetes (PWV: -3.5 % \[CI: -4.8; -2.1)\] $^{*^{p-value< 0.1}}$ CD: 4.8 % \[CI:1.3; 8.4\]$^{*^{p-value< 0.1}}$). For the indices SDNN and SDANN, the association with both cf-PWV and CD was significantly modified by both prediabetes and type 2 diabetes.

The association between HRV frequency-domain Z-score and cf-PWV was significantly modified from normal glucose metabolism by prediabetes (-5.7 %\[CI:-7.4; -3.9\]$^{*^{p-value< 0.01}}$) and type 2 diabetes (-3.9 %\[CI:-5.4; -2.3\]$^{*^{p-value< 0.05}}$) while CD was only modified by prediabetes (8.3 %\[CI:3.6; 13.2\]$^{*^{p-value< 0.01}}$) but not by type 2 diabetes (5.3 %\[CI:1.4; 9.4\]$^{*^{p-value< 0.1}}$). For the indices total power and ULF, the association with both cf-PWV and CD was significantly modified by both prediabetes and type 2 diabetes. Mean inter beat interval association with cf-PWV or CD was not significantly modified by diabetes status.

As we did not observe a stepwise increase in the modification of glucose metabolism status from prediabetes to type 2 diabetes, we excluded the subgroup with type 2 diabetes to test whether the association was gradually modified by dysglycemia. In this subgroup, the association between HRV time and frequency domain Z-scores and measures of arterial stiffness was modified by HbA1c (range of interaction p-values: 0.1 to 0.005) (see Figure x). For example, per SD lower HRV frequency domain Z-score at HbA1c 6.4% was associated with a 5.4% higher (CI: 3.5; 7.2) cf-PWV, which was 2.0% to 4.0% higher compared to the magnitude of association at HbA1c levels of 5.6% and 4.8% (see Figure xB). In CD, per SD lower HRV frequency domain Z-score at HbA1c 6.4% was associated with an 8.1% lower (CI: -13.5; -2.9) CD, which was 4.8% to 9.5% lower compared to the magnitude of association at HbA1c levels of 5.6% and 4.8% (see Figure xD). No association between HRV frequency domain Z-score and CD was observed at HbA1c levels between 4.8% and 5.2%.

::: {.center}
![](images/em_hba1c.pdf){width=4in}
:::

## Study II

### Descriptive

In ADDITION-PRO population consisted of 1,627 participant with a least 48-hour HRV measures, while 1,432 had all hour represented with hourly HRV and physical acceleration. The study population included different tiers of diabetes risk: 154 individuals at low risk (9%), 889 at high risk (51%), 314 with impaired fasting glucose (IFG) (18%), 226 with impaired glucose tolerance (IGT) (13%), and 161 with both IFG and IGT (9%). We splitted SDNN into categories by very-low (SDNN< 100 ms), low (SDNN 100-120 ms), middle (SDNN 121-140 ms), high (SDNN 141-160 ms) and very-high (SDNN >160 ms).

Participants in the lowest SDNN group (<100 ms) were older (67.4 ± 6.9 years), had higher BMI (28.1 ± 5.4), HbA1c (5.9 ± 0.9), triglycerides (1.5 ± 0.9 mmol/L), and resting heart rate (67.8 ± 5.7 bpm), were more likely to use anti-hypertensive medication (61%), and had lower physical activity energy expenditure (46.8 ± 24.0 kJ/day) compared to those with higher SDNN levels.

```{r}
#| echo: false
#| warning: false
#| message: false

table_add_pro <- readxl::read_xlsx(here("tables/table_1_dis.xlsx"))

table_add_pro[-nrow(table_add_pro), ] %>% 
  # dplyr::filter(!`**Characteristic**` %in% c("Ethnicity",
  #                                            "White",
  #                                            "Non-white",
  #                                            "Education",
  #                                            "Low (No education, (un)completed primary education, or lower vocational education)",
  #                                            "Middle (intermediate vocational education or higher secondary education)",
  #                                            "High (Higher vocational education or university education)",
  #                                            "None", 
  #                                            "Alcohol consumption",
  #                                            "Low (Women: ≤ 7, Men: ≤ 14)", 
  #                                            "High (Women: > 7, Men: > 14)",
  #                                            "Low (Women: â‰¤ 7, Men: â‰¤ 14)",
  #                                            "Smoking status", "Never", 
  #                                            "Former (quit > 6 months ago)", 
  #                                            "Former (quit < 6 months ago)", 
  #                                            "Current",  
  #                                            "ECGvar_LFHF", "ECGvar_NN50", 
  #                                            "ECGvar_SDSD", "med_DM")) %>%
      mutate_all(~ ifelse(is.na(.), "", .)) %>% 
  knitr::kable() %>%
  kableExtra::kable_classic() %>%
  kableExtra::kable_styling(latex_options = "scale_down")
```

### multiday HRV and MACE, heart failure, and all-cause mortality.

The mean multiday SDNN was 139.0 (32.3) ms, and the mean heart rate was 73.5 (9.1) bpm. In the fully adjusted model, SDNN per SD was associated with a lower incidence rate ratio (IRR) for MACE 0.82 (CI: 0.69; 0.97), heart failure 0.76 (CI: 0.58; 0.99), and mortality rate ratio of 0.79 (CI: 0.66; 0.94). When we pre-adjusted for resting heart rate, the proportion of the association explained between HRV and MACE, HF, and all-cause mortality was 14%, 25%, and 19%, respectively. We included knots in the model, which showed that the risk increased when SDNN fell below 120 to 110 ms (approximately below the 20th percentile), suggesting a potential cut-off point for higher risk. We therefore calculated the incidence rate (IR) at SDNN levels of 100 ms, 120 ms, and 160 ms, respectively, and plotted these as a function of age.
::: {.center}
![Multiday SDNN levels (100 ms, 120 ms, 160 ms) by age-specific incidence rates for A) major adverse cardiovascular events, B) heart failure, and C) all-cause mortality](images/addition_pro_hrv_ir_mace.pdf){width=4in}
:::

At age 65, the IR per 1000 person-years for MACE was 44.2 (CI: 33.5; 58.3) at SDNN = 100 ms, which was higher than the rates observed at SDNN = 120 ms (IR: 37.6 [CI: 29.2; 48.3]) and SDNN = 160 ms (IR: 34.7 [CI: 27.0; 44.5]) (Figure xA). The IR became higher with age, reaching its peak at age 72. For heart failure at age 65, the IR was 20.5 (CI: 15.3; 27.5) at SDNN = 100 ms, slightly higher than at SDNN = 120 ms (IR: 19.6 [CI: 14.8; 25.9]) and SDNN = 160 ms (IR: 19.8 [CI: 15.0; 26.2]) (Figure xB). The IR remained stable until age 70, after which it became higher. For all-cause mortality at age 65, the IR was 11.6 (CI: 6.3; 21.4) at SDNN = 100 ms, higher than at SDNN = 120 ms (IR: 7.6 [CI: 4.3; 13.3]) and SDNN = 160 ms (IR: 6.0 [CI: 3.4; 10.4]) (Figure xC). The IR for all-cause mortality became higher with age.

### Hourly HRV and MACE, heart failure, and all-cause mortality.

From the hourly recordings, we observed a clear periodicity in SDNN, mean heart rate (HR), sleep patterns, and physical acceleration. SDNN increased from 5–6 AM, peaking at 8–9 AM \[mean (SD)\], followed by a gradual decline, reaching its lowest point around 1 AM the next day \[mean SDNN (SD)\]. A similar circadian pattern was observed in heart rate, though its peak occurred two hour later at 10–11 AM. After peaking, heart rate remained stable throughout the afternoon before gradually decreasing.

::: {.center}
![Hourly SDNN levels (100 ms, 120 ms, 160 ms) by age-specific incidence rates for A) major adverse cardiovascular events, B) heart failure, and C) all-cause mortality. Figure adapted from [authors]. ---. (Paper II, appendix)](images/figure_ADD_PRO_risk_by_hour.pdf){width=7in}
:::


## Study III

### Patients characteristics

In study III, 179 participants with type 2 had measures of NT-proBNP and performed the CART test. CAN was present in 30% (n = 54) of participants (36% among those with valid CAN measurements). Meanwhile, 24% (n = 43) were unable to complete the CART assessment adequately, primarily due to irregular heart rhythms (n = 8) or insufficient air pressure during the Valsalva manoeuvre (n = 21). Compare to those without CAN, the participants with CAN were more women (41 % vs 33 %), were more sedentary (45% vs 36%), had a higher proportion with prior major CVD (41% vs 20%) and declined eGFR (\< 60) (36% vs 22%), higher levels of triglyceride (median 2.05 mmol/L vs 1.95 mmol/ L), were slightly older (median 62 years vs 61 years), had longer duration of type 2 diabetes (median 19 years vs 15 years), and higher use SGTL2-inhibitors (65% vs 60%) but lower use of GLP-1 RA (63% vs 70%). No other difference in clinical characteristic was observed.

### Relationship between CAN and elevated NT-proBNP

A greater proportion of individuals with CAN exhibited elevated NT-proBNP levels (\>125 pg/ml) (51.9%, n=52/78) compared to those without CAN (23.2%, n=26/112). The fully adjusted odds ratio (OR) for elevated NT-proBNP in individuals with CAN was 5.69 (95% CI: 1.95, 18.49) relative to those without CAN. Among the cardiovascular autonomic reflex tests (CART), the Valsalva maneuver demonstrated the strongest association with NT-proBNP (OR 9.00, 95% CI: 2.88, 33.09; n=51/75), followed by deep breathing (OR 3.30, 95% CI: 1.17, 9.77; n=33/133) and orthostatic hypertension (OR 4.04, 95% CI: 1.27, 13.77; n=24/146). No significant association was identified for the lying-to-standing test (OR 0.80, 95% CI: 0.32, 1.97; n=54/108). After imputing missing CART data, the OR for CAN in relation to elevated NT-proBNP declined to 2.94 (95% CI: 1.37, 6.56). Sensitivity analyses, which excluded participants using beta-blockers or those with a history of CVD, resulted in a smaller sample size and wider confidence intervals, though the overall association remained unchanged.

::: {layout-ncol="1"}
![Relationship betweeen CAN and indicators for heart failure. Figure from [authors]. Cardiovascular autonomic neuropathy and indices of heart failure in type 2 diabetes: The CANCAN Study. (Paper III, appendix)](images/heart_failure_distribution_can_figure_2.pdf)
:::
```

